---
title: 'R Workshop: Data Frames and Programming Structure'
author: "Tessa Chen"
date: "February 20, 2016"
output:
  ioslides_presentation:
    highlight: pygments
    widescreen: yes
  beamer_presentation:
    highlight: pygments
  slidy_presentation:
    highlight: pygments
---


## Outline

- Creating Data Frames
- Matrix-Like Operations
- Merging Data Frames
- Applying Functions to Data Frames
- Programming Structure
- Where to get help
<br>

All files we need in this workshop can be found: <br>
https://github.com/Ying-Ju/Miami-DataFest-Workshops

# Creating Data Frames
## Example 

```{r}
names = c('David', 'John', 'Mary')
quiz.1 = c(89, 93, 85)
quiz.2 = c(91, 88, 90)
Grade = data.frame(names, quiz.1, quiz.2, stringsAsFactors = TRUE)
Grade
```

## Example 

```{r}
Grade[[1]]
Grade$quiz.1
Grade[,3]
```

## Example (Reading a csv file)

```{r}
mypath <-'https://raw.githubusercontent.com/Ying-Ju/Miami-DataFest-Workshops/master/'
data1 <- paste(mypath,'tuition.csv', sep='')
tuition <- read.table(data1, header=TRUE, sep=",", as.is=FALSE)
dim(tuition)
names(tuition)
```

## Example (Filtering)

```{r}
tuition[tuition$Public.UO >= 8000,]
subset(tuition, Public.UO >= 8000)
```

## Example (Dealing with missing values)

```{r}
newtuition = rbind(tuition[1:3,], c(12131, NA, 5331))
complete = newtuition[complete.cases(newtuition),]
mean(newtuition$Public.UI)
mean(newtuition$Public.UI, na.rm=TRUE)
```
# Matrix-Like Operations
## Example (Applying apply())

```{r}
apply(tuition, 2, median)
apply(tuition, 2, mean)
```

## Example (Merging Data Frames)

```{r}
Grade1 = data.frame(students = c('David', 'Gabby', 'Mary'), 
                    quiz.3=c(88, 92, 85), 
                    stringsAsFactors=TRUE)
merge(Grade, Grade1, by.x="names", by.y="students")
lapply(Grade[,2:3], mean)
```

## Example (Applying tapply())
```{r}
Grade.example <- data.frame(
student <- 1:100,
grade <- rnorm(100, mean=70, sd=15),
gender <- gl(2, 50, labels=c("Male", "Female")))
summary(Grade.example)
```
- In this data set, we generate 100 students' grades by using the normal distribution with mean 70 and the standard deviation 15
- gl is a function to generate factor levels

## Example

```{r}
grade.mean = tapply(Grade.example$grade, Grade.example$gender, mean)
grade.sd = tapply(Grade.example$grade, Grade.example$gender, sd)
grade.quantile = tapply(Grade.example$grade, Grade.example$gender, 
                        function(x) quantile(x, c(0.05, 0.95)))
CI.left = c(grade.quantile$Male[1], grade.quantile$Female[1])
CI.right = c(grade.quantile$Male[2], grade.quantile$Female[2])
cbind(grade.mean, grade.sd, CI.left, CI.right)
```

## Applying Logistic Regression Models

$$P(Y=1|X=t)=\frac{1}{1+\exp[-(\beta_0+\beta_1t)]}$$

```{r}
aba <- read.table(paste(mypath,'ABALONE_data.txt', sep=''), sep = ",", header=T)
abamf <- aba[aba$Gender != "I",]
lftn <- function(clmn){
  glm(abamf$Gender~clmn, family = binomial)$coef
}
loall <- sapply(abamf[,-1],lftn)
loall
```


## How to write a function in R
-Example: $f(x,y,z)=a\cdot x+b \cdot y+c$

```{r}
myfunction <- function(x, y, c){
  c[1]*x+c[2]*y+c[3]
}
myfunction(2, 1, c(4, 1, 9))
```
- Use the comment: rm(list=ls())  to clean the workspace
- R takes only the last line of the function to be the return value 

# Programming Structure in R

## Programming Structure in R

- for
- while
- ifelse statment 
- if statement
- repeat

## Example: for loop
- Find the sum of $1+2+\cdots+n$. <br>
How to start?

```{r}
n <- 100
total.loop <- 0
for (i in 1:n) total.loop <- total.loop + i 
total.loop
```

- **Try**: Put the loop in a function with an input value $n$ and return the sum. 

## Example: while loop
- Find the sum of $1+2+\cdots+n$.

```{r}
n <- 100
total.while <- 0
k <- 0
while (k<n){
  k <- k + 1
  total.while <- total.while + k
}
total.while
```

- **Question**: What's the difference between if and while loops? When should we use if / while statments?

## Example: if statement

- Write a simple loop to find all even numbers and odd numbers in [15, 30] and print the statment "n is an even / odd number." 
```{r, eval=FALSE}
for (i in seq(15, 30)){
  if (i %% 2 == 0){
  print(paste(i, ' is an even number.'))
  }else{
    print(paste(i, ' is an odd number.'))
  }
}
```

## Example: ifelse statment

```{r, eval=FALSE}
  mynumber <- seq(15, 30)
  evenorodd <- ifelse(mynumber%%2==0, " is an even number.", " is an odd number.")
  for (i in 1:length(evenorodd)){
    print(paste(mynumber[i], evenorodd[i]))
  }
```

- **Question**: When should we use if / ifelse statements?

## Example: repeat
- Find the sum of $1+2+\cdots+n$.
```{r}
x <- 0
total.repeat <- 0
repeat {
   x <- x+1
   total.repeat <- total.repeat + x
   if (x == 100){
       break
   }
}
total.repeat
```

## Can a function call itself?
Yes! It is called a **recursive function**. (A function with a function) <br>
Example: Compute $1\times 2+ 2\times 3 + \ldots + n\times (n+1)$.

```{r}
Myfun_rec <- function(n){
  if (n==1){
    return(2)
  } else {
    return(n*(n+1)+Myfun_rec(n-1))
  }
}
```

## An example for computing fractals from iterated function systems
- http://krazydad.com/bestiary/bestiary_fern.html

```{r, echo=FALSE}
a = c(0, 0.85, 0.2, -0.15)
b = c(0, 0.04, -0.26, 0.28)
c = c(0, -0.04, 0.23, 0.26)
d = c(0.16, 0.85, 0.22, 0.24)
e = c(0, 0, 0, 0)
f = c(0, 1.6, 1.6, 0.44)

numits = 2000 # number of iterations
x = 0
y = 0
par(bg="black") # change the color for background 
plot(seq(-1, 10, by = 0.1), seq(-1, 10, by = 0.1), type = "n", 
     main = "fractal fern")
for (n in seq(1,numits)){
  k = sample(1:4, size = 1, replace = TRUE, prob = c(0.01, 0.85, 0.07, 0.07))
  newx = a[k]*x+b[k]*y+e[k]
  newy = c[k]*x+d[k]*y+f[k]
  x = newx
  y = newy
  if (n>10){
    points(x+3,y, col = "green", cex=0.5, pch=20)
    #points(x+5,y, col = "green", cex=0.5, pch=20)
  }
}
```

## Where to get help ?

1. To see documentation on any **function** in R, execute `?data.frame`, `?apply`, etc.
2. Go: http://www.orgs.miamioh.edu/datafest/index.html/resource.html
3. Ask people if you can!
4. Google it! (Better way to learn coding!)
